### 2018-03-09 submission/manuscript.Rmd
change : to . for titles

***Lower bacterial diversity is associated with increased odds ratio (OR) of tumors:***
We first assessed whether variation in broad community metrics like total number of operational taxonomic units (OTUs) (i.e. richness), the evenness of their abundance, and the overall diversity was associated with disease stage after controlling for study and variable region differences. In stool, there was a significant decrease in both evenness and diversity as disease severity progressed from normal to adenoma to carcinoma (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))` and `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`, respectively) [Figure 1]; there was not a significant difference for richness (P-value = `r paste(format(filter(stool_lme, alpha_metric == "sobs")[, "BH"], digits = 2, nsmall = 2))`). We next tested whether the decrease in these community metrics translated into significant ORs for having an adenoma or carcinoma. For fecal samples, the ORs for richness were not significantly greater than 1.0 for adenoma or carcinoma (P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "sobs")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2A]. The ORs for evenness were significantly higher than 1.0 for adenoma (OR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) and carcinoma (OR =  `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2B]. The ORs for diversity were only significantly greater than 1.0 for carcinoma (OR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))`), but not for adenoma (P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2C]. Although these OR are significantly greater than 1.0, it is doubtful that these are clinically meaningful values.

@@

***Disease progression is associated with community-wide changes in composition and abundance:*** Based on the differences in evenness and diversity, we next asked whether there were community-wide differences in the structure of the communities associated with different disease stages. We identified significant bacterial community differences in the stool of patients with adenomas relative to those with normal colons in `r length(as.data.frame(filter(adn_beta_stool, pvalue < 0.05))[, "study"])` of `r length(as.data.frame(adn_beta_stool)[, "study"])` studies and in patients with carcinomas relative to those with normal colons in 6 of 7 studies (PERMANOVA; P-value < 0.05) [Table S2]. Similar to the analyses using stool samples, there were significant differences in bacterial community structure between subjects with normal colons and those with adenoma (1 of 2 studies) and carcinoma (1 of 3 studies) [Table S2]. For studies that used matched samples no differences in bacterial community structures were observed [Table S2]. Combined, these results indicate that there consistent and significant community-wide changes in the fecal community structure of subjects with carcinomas. However, the signal observed in subjects with adenomas or when using tissue samples was not as consistent. This is likely due to a smaller effect size or the relatively small sample sizes among the studies that characterized the tissue microbiota.

***Individual taxa are associated with significant ORs for carcinomas:*** 

@@

***Individual taxa with a significant OR do a poor job of differentiating subjects with normal colons and those with carcinoma:***
 We next asked whether those taxa that had a significant OR associated with having a normal colon or carcinomas could be used individually, to classify subjects as having a normal colon or carcinomas. OR values were caluclated based on whether the relative abundance for a taxon in a subject was above or below the median relative abundance for that taxon across all subjects in a study. To measure the ability of these taxa to classify individuals we instead generated receiver operator characteristic (ROC) curves for each taxon in each study and calculated the area under the curve (AUC). This allowed us to use a more fluid relative abundance threshold for defining disease status. Using data from stool samples, the `r length(crc_OR_stool$measure)` taxa did no better at classifying the subjects than one would expect by chance (i.e. AUC=0.50) [Figure 3A]. The taxa that performed the best included *Clostridium XI*, *Ruminococcus*, and *Escherichia*, however, these had median AUC values less than `r round(max(filter(ind_crc_combined_stool_auc, taxa %in% c("Escherichia", "Clostridium_XI", "Ruminococcus"))[, "auc"]), digits = 3)`. Likewise, in unmatched tissue samples the `r length(crc_OR_stool$measure)` taxa with significant ORs taxa were marginally better than one would expect by chance [Figure 3B]. The relative abundance of Dorea was the best predictor of carcinomas and its median AUC was only `r round(max(filter(ind_crc_combined_unmatched_auc, taxa == "Dorea")[, "auc"]), digits = 3)`. These results suggest that although these taxa are associated with a decreased or increased OR for the presences of carcinomas, individually, they do a poor job of classifying a subject’s disease status.

***Combined taxa model classifies subjects better than using individual taxa:*** 
Instead of attempting to classify subjects based on individual taxa, next we combined information from the individual taxa and evaluated the ability to classify a subject’s disease status using Random Forest models. For data from stool samples, the combined model had an AUC of `r round(median(filter(crc_stool_model_aucs, study == train_model, model_type == "full_model")[, "AUC"]), digits = 2)`, which was significantly higher than any of the AUC values for the individual taxa (P-value < `r round(max(stool_ind_vs_full_taxa$pvalue), digits = 3)`). For the full taxa models using stool, *Bacteroides* and *Lachnospiraceae* were the most common taxa in the top 10% mean decrease in accuracy (MDA) across studies [Figure S2]. Of the `r length(crc_unmatched_OR_tissue$measure)` taxa with significant ORs, all `r length(crc_unmatched_OR_tissue$measure)` were among the top 10% most important taxa as measured by mean decrease in accuracy, in at least one study. The most important taxa across study within the significant OR taxa only models for stool were *Ruminococcus* and *Clostridium XI* [Figure 5A]. Similarly, using data from the unmatched tissue samples, the combined model had an AUC of `r round(median(filter(crc_tissue_model_aucs, study == train_model, model_type == "full_model", type == "unmatched")[, "AUC"]), digits = 2)`, which was significantly higher than the AUC values for Blautia and Weissella (P-value < `r round(max(filter(unmatched_tissue_ind_vs_full_taxa, taxa != "Dorea")[, "pvalue"]), digits = 3)`).  For the full taxa models using unmatched tissue, *Lachnospiraceae*, *Bacteroidaceae*, and *Ruminococcaceae* were the most common taxa in the top 10% mean decrease in accuracy across studies [Figure S3]. For the singificant OR taxa unmatched tissue models both *Dorea* and *Blautia* were the important based on mean decrease in accuracy [Figure 5B]. Pooling the information from the taxa with significant ORs results in a model that outperforms classifications made using individual taxa.

***Performance of models based on taxa relative abundance in full community are better than those based on taxa with significant ORs:*** 
Next, we asked whether a Random Forest classification model built using all of the taxa found in the communities would outperform the models generated using those taxa with a significant OR. Similar to our inability to identify taxa associated with a significant OR for the presence of adenomas, the median AUCs to classify subjects as having normal colons or having adenomas using data from stool or tissue samples were only marginally better than 0.5 for any study (median AUC = `r median(c(filter(adn_stool_model_aucs, study == train_model)[, "AUC"], filter(adn_tissue_model_aucs, study == train_model)[, "AUC"]), na.rm = T)` [`r min(c(filter(adn_stool_model_aucs, study == train_model)[, "AUC"], filter(adn_tissue_model_aucs, study == train_model)[, "AUC"]), na.rm = T)` - `r max(c(filter(adn_stool_model_aucs, study == train_model)[, "AUC"], filter(adn_tissue_model_aucs, study == train_model)[, "AUC"]), na.rm = T)`]) [Figure 4A & S4A]. In contrast, the models for classifying subjects as having normal colons or having carcinomas using data from stool or tissue samples yielded AUC values meaningfully higher than 0.5 [Figure 4B & S4B-C]. When we compared the models based on all of the taxa in a community to models based on the taxa with significant ORs, the results were mixed. Using the data from stool samples we found that the AUC for 6 of 7 studies increased by an average of `r round(mean((1 - (filter(crc_stool_model_aucs, study == train_model, model_type == "full_model", study != "flemer")[, "AUC"] / filter(crc_stool_model_aucs, study == train_model, model_type == "select_model", study != "flemer")[, "AUC"]))*100), digits = 2)`%) and AUC for the Flemer study decreased by `r round(abs(mean(((1 - filter(crc_stool_model_aucs, study == train_model, model_type == "full_model", study == "flemer")[, "AUC"] / filter(crc_stool_model_aucs, study == train_model, model_type == "select_model", study == "flemer")[, "AUC"]))*100)), digits = 2)`%). The overall improvement in performance was statistically significant (mean = `r abs(round(mean((1 - (filter(crc_stool_model_aucs, study == train_model, model_type == "full_model")[, "AUC"] / filter(crc_stool_model_aucs, study == train_model, model_type == "select_model")[, "AUC"]))*100), digits = 2))`%, one-tailed paired T-test; P-value = `r round(t.test(filter(crc_stool_model_aucs, study == train_model, model_type == "full_model")[, "AUC"], filter(crc_stool_model_aucs, study == train_model, model_type == "select_model")[, "AUC"], alternative = "greater", paired = TRUE)$p.value, digits = 3)`). Similarly, using the data from unmatched tissue samples we found that the AUC of studies increased by an average of `r round(mean((filter(crc_tissue_model_aucs, study == train_model, model_type == "full_model", type == "unmatched")[, "AUC"] - filter(crc_tissue_model_aucs, study == train_model, model_type == "select_model", type == "unmatched")[, "AUC"])*100), digits = 2)`% when we used all of the taxa rather than the 3 taxa with significant ORs (one-tailed paired T-test; P-value = `r round(t.test(filter(crc_tissue_model_aucs, study == train_model, model_type == "full_model", type == "unmatched")[, "AUC"], filter(crc_tissue_model_aucs, study == train_model, model_type == "select_model", type == "unmatched")[, "AUC"], alternative = "greater", paired = TRUE)$p.value, digits = 3)`). Although the significant OR taxa models can classify indidivuals with and without carcinoma tumors, they are still missing taxa from the full community models that can increase the model accuracy.

***Performance of models based on OTU relative abundance in full community are not significantly better than those based on taxa with significant ORs:*** 
The previous models were based on relative abundance data where sequences were classified to coarse taxonomic assignments (i.e. typically genus or family level). To determine whether model performance improved with finer scale classification, we assigned sequences to operational taxonomic units (OTUs) where the similarity among sequences within an OTU was more than 97%. We again found that classification models built using all of the sequence data for a community did a poor job of differentiating between subjects with normal colons and those with adenomas (median AUC: `r round(median(filter(adn_stool_model_aucs, model_type == "full_otu")[, "AUC"]), digits = 2)` [`r round(min(filter(adn_stool_model_aucs, model_type == "full_otu")[, "AUC"]), digits = 2)`- `r round(max(filter(adn_stool_model_aucs, model_type == "full_otu")[, "AUC"]), digits = 2)`]), but did a good job of differentiating between subjects with normal colons and those with carcinomas (median AUC: `r round(median(filter(crc_stool_model_aucs, model_type == "full_otu")[, "AUC"]), digits = 2)` [`r round(min(filter(crc_stool_model_aucs, model_type == "full_otu")[, "AUC"]), digits = 2)`- `r round(max(filter(crc_stool_model_aucs, model_type == "full_otu")[, "AUC"]), digits = 2)`]). The OTU-based models performed similarly to those constructed using the taxa with significant ORs (one-tailed paired T-test; P-value = `r round(t.test(filter(crc_tissue_model_aucs, study == train_model, model_type == "select_model", type == "unmatched")[, "AUC"], filter(crc_tissue_model_aucs, study == train_model, model_type == "full_otu", type == "unmatched")[, "AUC"], alternative = "greater", paired = TRUE)$p.value, digits = 3)`) and those using all taxa (one-tailed paired T-test; P-value = `r round(t.test(filter(crc_tissue_model_aucs, study == train_model, model_type == "full_model", type == "unmatched")[, "AUC"], filter(crc_tissue_model_aucs, study == train_model, model_type == "full_otu", type == "unmatched")[, "AUC"], alternative = "greater", paired = TRUE)$p.value, digits = 3)`) [Figure 4]. Among the OTUs that had the highest mean decrease in accuracy for the OTU-based models, we found that OTUs that affiliated with all of the `r length(crc_OR_stool$measure)` taxa that had a significant OR were within the top 10% for at least one study. This result was surprising as it indicated that a finer scale classification of the sequences and thus a larger number of features to select from, did not yield improved classification of the subjects.

***Generalizability of taxon-based models trained on one dataset to the other datasets:*** 
Considering the good performance of the Random Forest models using taxa with a significant OR and using all of the taxa, we next asked how well the models would perform when given data from a different subject cohort. For instance, if a model was trained using data from the Ahn study, we wanted to know how well it would perform using the data from the Baxter study. We found the models trained using the taxa with a significant OR all had a higher median AUC than the models trained using all of the taxa when tested on the other datasets [Figure 6 & S5]. As might be expected, the difference between the performance of the modelling approaches appeared to vary with the size of the training cohort (R^2^ = `r round(lm(auc_median ~ totals, data = correlation_data)$coefficients[1], digits = 2)`) [Figure 6]. These data suggest that given a sufficient number of subjects with normal colons and carcinomas, Random Forest models trained using a small number of taxa can accurately classify individuals from a different cohort.

@@

***Datasets:***
The studies used for this meta-analysis were identified through the review articles written by Keku, et al. [@keku_gastrointestinal_2015] and Vogtmann, et al. [@vogtmann_epidemiologic_2016]. Additional studies, not mentioned in those reviews were obtained based on the authors’ knowledge of the literature. Studies were included that used tissue or feces as their sample source for 454 or Illumina 16S rRNA gene sequencing. Some studies (N = 12) were excluded because they did not have publicly available sequences or did not have metadata in which the authors were able to share. We were able to obtain sequence data and metadata from the following studies: Ahn, *et al.* [@ahn_human_2013], Baxter, *et al.* [@baxter_microbiota-based_2016], Brim, *et al.* [@brim_microbiome_2013], Burns, *et al.* [@burns_virulence_2015], Chen, *et al.*  [@chen_human_2012], Dejea, *et al.* [@dejea_microbiota_2014], Flemer, *et al.* [@flemer_tumour-associated_2017], Geng, *et al.* [@geng_diversified_2013], Hale, *et al.* [@hale_shifts_2017], Kostic, *et al.* [@kostic_genomic_2012], Lu, *et al.* [@lu_mucosal_2016], Sanapareddy, *et al.* [@sanapareddy_increased_2012], Wang, *et al.* [@wang_structural_2012], Weir, *et al.* [@weir_stool_2013], and Zeller, *et al.*  [@zeller_potential_2014]. The Zackular [@zackular_human_2014] study was excluded because their 90 individuals were contained within the larger Baxter study [@baxter_microbiota-based_2016]. The Kostic study was excluded because after we processed the sequences, all of the case samples had 100 or fewer sequences. The final analysis included 14 studies (Tables 1 and 2). There were seven studies with only fecal samples (Ahn, Baxter, Brim, Hale, Wang, Weir, and Zeller), five studies with only tissue samples (Burns, Dejea, Geng, Lu, Sanapareddy), and two studies with both fecal and tissue samples (Chen and Flemer). After curating the sequences, `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)` stool samples and `r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` tissue samples remained in the analysis [Tables 1 and 2].

***Sequence Processing:***
Raw sequence data and metadata were primarily obtained from the Sequence Read Archive (SRA) and dbGaP. Other sequence and metadata were obtained directly from the authors (n = 4, [@flemer_tumour-associated_2017; @hale_shifts_2017; @sanapareddy_increased_2012; @weir_stool_2013]). Each dataset was processed separately using mothur (v1.39.3) [@schloss_introducing_2009] using the default quality filtering methods for both 454 and Illumina sequence data. If it was not possible to use the defaults because the sequences were trimmed too much, then the stated quality cut-offs from the original study were used. Chimeric sequences were identified and removed using VSEARCH [@rognes_vsearch_2016]. The curated sequences were assigned to OTUs at 97% similarity using the OptiClust algorithm [@westcott_opticlust_2017] and classified to the deepest taxonomic level that had 80% support using the naïve Bayesian classifier trained on the RDP taxonomy outline (version 14, [@rdp_Wang2007]).

>>>>

***Lower bacterial diversity is associated with increased odds ratio (OR) of tumors.***
We first assessed whether variation in broad community metrics like total number of operational taxonomic units (OTUs) (i.e. richness), the evenness of their abundance, and the overall diversity was associated with disease stage after controlling for study and variable region differences. In stool, there was a significant decrease in both evenness and diversity as disease severity progressed from normal to adenoma to carcinoma (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))` and `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`, respectively) [Figure 1]; there was not a significant difference for richness (P-value = `r paste(format(filter(stool_lme, alpha_metric == "sobs")[, "BH"], digits = 2, nsmall = 2))`). We next tested whether the decrease in these community metrics translated into significant ORs for having an adenoma or carcinoma. For fecal samples, the ORs for richness were not significantly greater than 1.0 for adenoma or carcinoma (P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "sobs")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2A]. The ORs for evenness were significantly higher than 1.0 for adenoma (OR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) and carcinoma (OR =  `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2B]. The ORs for diversity were only significantly greater than 1.0 for carcinoma (OR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))`), but not for adenoma (P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2C]. Although these OR are significantly greater than 1.0, it is doubtful that these are clinically meaningful values.

@@

***Disease progression is associated with community-wide changes in composition and abundance.*** Based on the differences in evenness and diversity, we next asked whether there were community-wide differences in the structure of the communities associated with different disease stages. We identified significant bacterial community differences in the stool of patients with adenomas relative to those with normal colons in `r length(as.data.frame(filter(adn_beta_stool, pvalue < 0.05))[, "study"])` of `r length(as.data.frame(adn_beta_stool)[, "study"])` studies and in patients with carcinomas relative to those with normal colons in 6 of 7 studies (PERMANOVA; P-value < 0.05) [Table S2]. Similar to the analyses using stool samples, there were significant differences in bacterial community structure between subjects with normal colons and those with adenoma (1 of 2 studies) and carcinoma (1 of 3 studies) [Table S2]. For studies that used matched samples no differences in bacterial community structures were observed [Table S2]. Combined, these results indicate that there consistent and significant community-wide changes in the fecal community structure of subjects with carcinomas. However, the signal observed in subjects with adenomas or when using tissue samples was not as consistent. This is likely due to a smaller effect size or the relatively small sample sizes among the studies that characterized the tissue microbiota.

***Individual taxa are associated with significant ORs for carcinomas.*** 

@@

***Individual taxa with a significant OR do a poor job of differentiating subjects with normal colons and those with carcinoma.***
 We next asked whether those taxa that had a significant OR associated with having a normal colon or carcinomas could be used individually, to classify subjects as having a normal colon or carcinomas. OR values were caluclated based on whether the relative abundance for a taxon in a subject was above or below the median relative abundance for that taxon across all subjects in a study. To measure the ability of these taxa to classify individuals we instead generated receiver operator characteristic (ROC) curves for each taxon in each study and calculated the area under the curve (AUC). This allowed us to use a more fluid relative abundance threshold for defining disease status. Using data from stool samples, the `r length(crc_OR_stool$measure)` taxa did no better at classifying the subjects than one would expect by chance (i.e. AUC=0.50) [Figure 3A]. The taxa that performed the best included *Clostridium XI*, *Ruminococcus*, and *Escherichia*, however, these had median AUC values less than `r round(max(filter(ind_crc_combined_stool_auc, taxa %in% c("Escherichia", "Clostridium_XI", "Ruminococcus"))[, "auc"]), digits = 3)`. Likewise, in unmatched tissue samples the `r length(crc_OR_stool$measure)` taxa with significant ORs taxa were marginally better than one would expect by chance [Figure 3B]. The relative abundance of Dorea was the best predictor of carcinomas and its median AUC was only `r round(max(filter(ind_crc_combined_unmatched_auc, taxa == "Dorea")[, "auc"]), digits = 3)`. These results suggest that although these taxa are associated with a decreased or increased OR for the presences of carcinomas, individually, they do a poor job of classifying a subject’s disease status.

***Combined taxa model classifies subjects better than using individual taxa.*** 
Instead of attempting to classify subjects based on individual taxa, next we combined information from the individual taxa and evaluated the ability to classify a subject’s disease status using Random Forest models. For data from stool samples, the combined model had an AUC of `r round(median(filter(crc_stool_model_aucs, study == train_model, model_type == "full_model")[, "AUC"]), digits = 2)`, which was significantly higher than any of the AUC values for the individual taxa (P-value < `r round(max(stool_ind_vs_full_taxa$pvalue), digits = 3)`). For the full taxa models using stool, *Bacteroides* and *Lachnospiraceae* were the most common taxa in the top 10% mean decrease in accuracy (MDA) across studies [Figure S2]. Of the `r length(crc_unmatched_OR_tissue$measure)` taxa with significant ORs, all `r length(crc_unmatched_OR_tissue$measure)` were among the top 10% most important taxa as measured by mean decrease in accuracy, in at least one study. The most important taxa across study within the significant OR taxa only models for stool were *Ruminococcus* and *Clostridium XI* [Figure 5A]. Similarly, using data from the unmatched tissue samples, the combined model had an AUC of `r round(median(filter(crc_tissue_model_aucs, study == train_model, model_type == "full_model", type == "unmatched")[, "AUC"]), digits = 2)`, which was significantly higher than the AUC values for Blautia and Weissella (P-value < `r round(max(filter(unmatched_tissue_ind_vs_full_taxa, taxa != "Dorea")[, "pvalue"]), digits = 3)`).  For the full taxa models using unmatched tissue, *Lachnospiraceae*, *Bacteroidaceae*, and *Ruminococcaceae* were the most common taxa in the top 10% mean decrease in accuracy across studies [Figure S3]. For the singificant OR taxa unmatched tissue models both *Dorea* and *Blautia* were the important based on mean decrease in accuracy [Figure 5B]. Pooling the information from the taxa with significant ORs results in a model that outperforms classifications made using individual taxa.

***Performance of models based on taxa relative abundance in full community are better than those based on taxa with significant ORs.*** 
Next, we asked whether a Random Forest classification model built using all of the taxa found in the communities would outperform the models generated using those taxa with a significant OR. Similar to our inability to identify taxa associated with a significant OR for the presence of adenomas, the median AUCs to classify subjects as having normal colons or having adenomas using data from stool or tissue samples were only marginally better than 0.5 for any study (median AUC = `r median(c(filter(adn_stool_model_aucs, study == train_model)[, "AUC"], filter(adn_tissue_model_aucs, study == train_model)[, "AUC"]), na.rm = T)` [`r min(c(filter(adn_stool_model_aucs, study == train_model)[, "AUC"], filter(adn_tissue_model_aucs, study == train_model)[, "AUC"]), na.rm = T)` - `r max(c(filter(adn_stool_model_aucs, study == train_model)[, "AUC"], filter(adn_tissue_model_aucs, study == train_model)[, "AUC"]), na.rm = T)`]) [Figure 4A & S4A]. In contrast, the models for classifying subjects as having normal colons or having carcinomas using data from stool or tissue samples yielded AUC values meaningfully higher than 0.5 [Figure 4B & S4B-C]. When we compared the models based on all of the taxa in a community to models based on the taxa with significant ORs, the results were mixed. Using the data from stool samples we found that the AUC for 6 of 7 studies increased by an average of `r round(mean((1 - (filter(crc_stool_model_aucs, study == train_model, model_type == "full_model", study != "flemer")[, "AUC"] / filter(crc_stool_model_aucs, study == train_model, model_type == "select_model", study != "flemer")[, "AUC"]))*100), digits = 2)`%) and AUC for the Flemer study decreased by `r round(abs(mean(((1 - filter(crc_stool_model_aucs, study == train_model, model_type == "full_model", study == "flemer")[, "AUC"] / filter(crc_stool_model_aucs, study == train_model, model_type == "select_model", study == "flemer")[, "AUC"]))*100)), digits = 2)`%). The overall improvement in performance was statistically significant (mean = `r abs(round(mean((1 - (filter(crc_stool_model_aucs, study == train_model, model_type == "full_model")[, "AUC"] / filter(crc_stool_model_aucs, study == train_model, model_type == "select_model")[, "AUC"]))*100), digits = 2))`%, one-tailed paired T-test; P-value = `r round(t.test(filter(crc_stool_model_aucs, study == train_model, model_type == "full_model")[, "AUC"], filter(crc_stool_model_aucs, study == train_model, model_type == "select_model")[, "AUC"], alternative = "greater", paired = TRUE)$p.value, digits = 3)`). Similarly, using the data from unmatched tissue samples we found that the AUC of studies increased by an average of `r round(mean((filter(crc_tissue_model_aucs, study == train_model, model_type == "full_model", type == "unmatched")[, "AUC"] - filter(crc_tissue_model_aucs, study == train_model, model_type == "select_model", type == "unmatched")[, "AUC"])*100), digits = 2)`% when we used all of the taxa rather than the 3 taxa with significant ORs (one-tailed paired T-test; P-value = `r round(t.test(filter(crc_tissue_model_aucs, study == train_model, model_type == "full_model", type == "unmatched")[, "AUC"], filter(crc_tissue_model_aucs, study == train_model, model_type == "select_model", type == "unmatched")[, "AUC"], alternative = "greater", paired = TRUE)$p.value, digits = 3)`). Although the significant OR taxa models can classify indidivuals with and without carcinoma tumors, they are still missing taxa from the full community models that can increase the model accuracy.

***Performance of models based on OTU relative abundance in full community are not significantly better than those based on taxa with significant ORs.*** 
The previous models were based on relative abundance data where sequences were classified to coarse taxonomic assignments (i.e. typically genus or family level). To determine whether model performance improved with finer scale classification, we assigned sequences to operational taxonomic units (OTUs) where the similarity among sequences within an OTU was more than 97%. We again found that classification models built using all of the sequence data for a community did a poor job of differentiating between subjects with normal colons and those with adenomas (median AUC: `r round(median(filter(adn_stool_model_aucs, model_type == "full_otu")[, "AUC"]), digits = 2)` [`r round(min(filter(adn_stool_model_aucs, model_type == "full_otu")[, "AUC"]), digits = 2)`- `r round(max(filter(adn_stool_model_aucs, model_type == "full_otu")[, "AUC"]), digits = 2)`]), but did a good job of differentiating between subjects with normal colons and those with carcinomas (median AUC: `r round(median(filter(crc_stool_model_aucs, model_type == "full_otu")[, "AUC"]), digits = 2)` [`r round(min(filter(crc_stool_model_aucs, model_type == "full_otu")[, "AUC"]), digits = 2)`- `r round(max(filter(crc_stool_model_aucs, model_type == "full_otu")[, "AUC"]), digits = 2)`]). The OTU-based models performed similarly to those constructed using the taxa with significant ORs (one-tailed paired T-test; P-value = `r round(t.test(filter(crc_tissue_model_aucs, study == train_model, model_type == "select_model", type == "unmatched")[, "AUC"], filter(crc_tissue_model_aucs, study == train_model, model_type == "full_otu", type == "unmatched")[, "AUC"], alternative = "greater", paired = TRUE)$p.value, digits = 3)`) and those using all taxa (one-tailed paired T-test; P-value = `r round(t.test(filter(crc_tissue_model_aucs, study == train_model, model_type == "full_model", type == "unmatched")[, "AUC"], filter(crc_tissue_model_aucs, study == train_model, model_type == "full_otu", type == "unmatched")[, "AUC"], alternative = "greater", paired = TRUE)$p.value, digits = 3)`) [Figure 4]. Among the OTUs that had the highest mean decrease in accuracy for the OTU-based models, we found that OTUs that affiliated with all of the `r length(crc_OR_stool$measure)` taxa that had a significant OR were within the top 10% for at least one study. This result was surprising as it indicated that a finer scale classification of the sequences and thus a larger number of features to select from, did not yield improved classification of the subjects.

***Generalizability of taxon-based models trained on one dataset to the other datasets.*** 
Considering the good performance of the Random Forest models using taxa with a significant R and using all of the taxa, we next asked how well the models would perform when given data from a different subject cohort. For instance, if a model was trained using data from the Ahn study, we wanted to know how well it would perform using the data from the Baxter study. We found the models trained using the taxa with a significant OR all had a higher median AUC than the models trained using all of the taxa when tested on the other datasets [Figure 6 & S5]. As might be expected, the difference between the performance of the modelling approaches appeared to vary with the size of the training cohort (R^2^ = `r round(lm(auc_median ~ totals, data = correlation_data)$coefficients[1], digits = 2)`) [Figure 6]. These data suggest that given a sufficient number of subjects with normal colons and carcinomas, Random Forest models trained using a small number of taxa can accurately classify individuals from a different cohort.

@@

***Datasets.***
The studies used for this meta-analysis were identified through the review articles written by Keku, et al. [@keku_gastrointestinal_2015] and Vogtmann, et al. [@vogtmann_epidemiologic_2016]. Additional studies, not mentioned in those reviews were obtained based on the authors’ knowledge of the literature. Studies were included that used tissue or feces as their sample source for 454 or Illumina 16S rRNA gene sequencing. Some studies (N = 12) were excluded because they did not have publicly available sequences or did not have metadata in which the authors were able to share. We were able to obtain sequence data and metadata from the following studies: Ahn, *et al.* [@ahn_human_2013], Baxter, *et al.* [@baxter_microbiota-based_2016], Brim, *et al.* [@brim_microbiome_2013], Burns, *et al.* [@burns_virulence_2015], Chen, *et al.*  [@chen_human_2012], Dejea, *et al.* [@dejea_microbiota_2014], Flemer, *et al.* [@flemer_tumour-associated_2017], Geng, *et al.* [@geng_diversified_2013], Hale, *et al.* [@hale_shifts_2017], Kostic, *et al.* [@kostic_genomic_2012], Lu, *et al.* [@lu_mucosal_2016], Sanapareddy, *et al.* [@sanapareddy_increased_2012], Wang, *et al.* [@wang_structural_2012], Weir, *et al.* [@weir_stool_2013], and Zeller, *et al.*  [@zeller_potential_2014]. The Zackular [@zackular_human_2014] study was excluded because their 90 individuals were contained within the larger Baxter study [@baxter_microbiota-based_2016]. The Kostic study was excluded because after we processed the sequences, all of the case samples had 100 or fewer sequences. The final analysis included 14 studies (Tables 1 and 2). There were seven studies with only fecal samples (Ahn, Baxter, Brim, Hale, Wang, Weir, and Zeller), five studies with only tissue samples (Burns, Dejea, Geng, Lu, Sanapareddy), and two studies with both fecal and tissue samples (Chen and Flemer). After curating the sequences, `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)` stool samples and `r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` tissue samples remained in the analysis [Tables 1 and 2].

***Sequence Processing.***
Raw sequence data and metadata were primarily obtained from the Sequence Read Archive (SRA) and dbGaP. Other sequence and metadata were obtained directly from the authors (n = 4, [@flemer_tumour-associated_2017; @hale_shifts_2017; @sanapareddy_increased_2012; @weir_stool_2013]). Each dataset was processed separately using mothur (v1.39.3) [@schloss_introducing_2009] using the default quality filtering methods for both 454 and Illumina sequence data. If it was not possible to use the defaults because the sequences were trimmed too much, then the stated quality cut-offs from the original study were used. Chimeric sequences were identified and removed using VSEARCH [@rognes_vsearch_2016]. The curated sequences were assigned to OTUs at 97% similarity using the OptiClust algorithm [@westcott_opticlust_2017] and classified to the deepest taxonomic level that had 80% support using the naïve Bayesian classifier trained on the RDP taxonomy outline (version 14, [@rdp_Wang2007]).
